---
title: "Ensemble consensus over the GRASP pool (LCDA-ECG)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ensemble consensus over the GRASP pool (LCDA-ECG)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7.4, fig.height = 4.2)
library(lcdaGRASP)
has_gg <- requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("dplyr", quietly = TRUE)
has_tidyr <- requireNamespace("tidyr", quietly = TRUE)
rec <- lcda_data("lcda_ecg"); ov <- lcda_data("overlap_lcda_ecg"); st <- lcda_data("stability_pool")
```

# Why throw the pool away?

A multi-start search produces a whole **pool** of diverse partitions, yet the
GRASP/Reactive variants keep only the single best-by-$(Q,H)$ solution. Following
Ensemble Clustering for Graphs (Poulin & Théberge, 2019), `lcda_ecg()` instead
aggregates the pool: the fraction of pool partitions in which two endpoints
share a community becomes an edge **co-association** weight, and the reweighted
graph is re-clustered for a **consensus partition**. Three by-products come for
free: a consensus leader, a node-confidence map, and overlapping communities.

```{r live}
g <- igraph::make_graph("Zachary")
res <- lcda_ecg(g, B = 32, overlap = TRUE, tau = 0.6, seed = 1)
res
cat("consensus leaders:", res$leaders, "\n")
cat("overlap (bridge) nodes:", which(res$is_overlap), "\n")
```

# Consensus closes the recovery gap

```{r recovery, eval = !is.null(rec) && has_gg}
library(ggplot2); library(dplyr)
rec$results$summary |>
  ggplot(aes(mu, NMI, colour = method, shape = method)) +
  geom_line(linewidth = 0.9) + geom_point(size = 2) +
  labs(title = "Recovery vs the planted LFR partition",
       subtitle = "LCDA-ECG matches ECG and outperforms Leiden (advantage at high mixing)",
       x = expression(mu), y = "NMI") +
  theme_minimal(base_size = 10) + theme(legend.position = "bottom")
```

The fixed/Reactive variants trail Leiden and ECG on recovery; the
ensemble-consensus variant **LCDA-ECG** matches ECG and overtakes Leiden in the
high-mixing regime, where consensus is most valuable. (It stays on par with ECG
rather than beating it.)

# Overlapping nodes show bridge-like connectivity

```{r overlap, eval = !is.null(ov) && has_gg && has_tidyr}
ov$results$per_graph |>
  group_by(mu) |>
  summarise(`overlap node` = mean(P_overlap, na.rm = TRUE),
            `non-overlap node` = mean(P_nonoverlap, na.rm = TRUE), .groups = "drop") |>
  tidyr::pivot_longer(-mu, names_to = "kind", values_to = "P") |>
  ggplot(aes(factor(mu), P, fill = kind)) +
  geom_col(position = position_dodge(0.7), width = 0.65) +
  labs(title = "Participation coefficient of overlap vs non-overlap nodes",
       x = expression(mu), y = "participation coefficient", fill = NULL) +
  theme_minimal(base_size = 10) + theme(legend.position = "bottom")
```

Nodes that LCDA-ECG places in more than one community have a significantly
higher participation coefficient (their edges spread across communities), i.e.
bridge-like connectivity. Because both the participation coefficient and the
overlap-assignment rule read the same node-to-community edge distribution, this
is a consistency check rather than an independent validation.

# The pool stabilises (so the pool size is principled)

```{r stability, eval = !is.null(st) && has_gg && has_tidyr}
st$results |>
  group_by(B) |>
  summarise(`co-assoc drift` = mean(coassoc_drift, na.rm = TRUE),
            `consensus stability` = mean(consensus_stability, na.rm = TRUE),
            `recovery` = mean(recovery), .groups = "drop") |>
  tidyr::pivot_longer(-B, names_to = "metric", values_to = "value") |>
  ggplot(aes(B, value, colour = metric)) +
  geom_line(linewidth = 0.9) + geom_point(size = 1.5) +
  scale_x_continuous(trans = "log2", breaks = c(2, 4, 8, 16, 32, 64, 128)) +
  labs(title = "Co-association converges as the pool grows",
       x = "pool size B (log2)", y = NULL, colour = NULL) +
  theme_minimal(base_size = 10) + theme(legend.position = "bottom")
```

As $B$ grows the co-association drift vanishes and both the consensus partition
and its recovery plateau by $B\approx 32$ --- a stability-based stopping rule,
consistent with the asymptotic concentration of the Reactive update.

# Reproducibility and data provenance

Every figure above is read from datasets shipped with the package; nothing is
simulated at build time. Each dataset records the package version that **generated** it (not
necessarily the version you installed: the generators are re-run only when the
algorithms change), the release it first shipped in, and a SHA-256 checksum
matching `inst/extdata/SHA256SUMS`:

```{r provenance}
do.call(rbind, lapply(c("lcda_ecg", "overlap_lcda_ecg", "stability_pool"),
                      lcda_provenance))
```

Regenerate with `data-raw/97_lcda_ecg.R`, `data-raw/99_overlap_lcda_ecg.R` and
`data-raw/100_stability_pool.R` (each fixes a single documented seed).

**Scope and honest reading.** LCDA-ECG is *competitive with* ECG — which remains
marginally ahead on pure recovery — and overtakes Leiden only at high mixing
($\mu = 0.5$). The overlapping-community result is a *consistency check*
(overlap nodes are structural bridges), not a validation against planted
overlap, which the disjoint LFR generator does not provide.

# References

- Poulin, V., & Théberge, F. (2019). Ensemble clustering for graphs.
  *Applied Network Science*, 4, 51.
  [doi:10.1007/s41109-019-0162-z](https://doi.org/10.1007/s41109-019-0162-z)
