---
title: "Getting started with lcdaGRASP"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with lcdaGRASP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
```

# Overview

`lcdaGRASP` implements the four algorithms of Ospina, Silva, Matos Junior,
Leite and Ochi (2026) for **joint community-and-leader detection**:

- `lcda_grasp()` --- fixed-parameter GRASP, variants 1 (static centrality)
  and 2 (adaptive centrality recomputed each iteration).
- `lcda_gr()` --- Reactive GRASP, variants 1 and 2, with a bivariate
  self-tuning mechanism over a pool of `(alpha_c, alpha_s)` pairs.

The performance-critical kernels (incremental modularity, community
affinities, similarity vectors, the VNMI multi-improvement extension, and
power-iteration eigenvector centrality) are implemented in C++ via Rcpp.
The package follows a tidyverse style: snake_case throughout, `tibble`
outputs where tabular, `cli` for logging, and a `verbose` flag on every
substantive function.

# Quick start

```{r quick}
library(lcdaGRASP)
library(igraph)

g <- make_graph("Zachary")           # 34-node karate club
res <- lcda_grasp(g,
                  alpha_c = 0.1, alpha_s = 0.3, B = 30,
                  centrality = "eigen", similarity = "hpi",
                  seed = 1)
print(res)
```

Set `verbose = TRUE` for a `cli` progress bar and a one-line summary:

```{r verbose}
invisible(lcda_grasp(g, B = 10, seed = 1, verbose = TRUE))
```

# Inspecting the GRASP trajectory

```{r traj}
plot_grasp_trajectory(res)
```

# Switching to the Reactive variant

```{r reactive}
res_r <- lcda_gr(g, variant = 1, B = 100, seed = 1)
print(res_r)
plot_reactive_pk(res_r)
```

The print method reports the fraction of iterations in which the
lexicographic tie-break on `H` (Node-Connection Entropy) was decisive ---
iterations where `Q` was tied with the running best and `H` made the
difference. In typical networks this fraction is small; see the
*Reactive convergence* and *Pool sensitivity* articles.

# Statistical comparison

The paper uses a Kruskal--Wallis omnibus test followed by pairwise
permutation tests with Bonferroni correction. `kruskal_then_permute()`
returns a tibble of pairwise contrasts and a compact-letter display:

```{r kw}
set.seed(1)
df <- data.frame(
  Q = c(rnorm(20, 0.40, 0.01), rnorm(20, 0.42, 0.01), rnorm(20, 0.42, 0.01)),
  cfg = rep(c("a", "b", "c"), each = 20)
)
out <- kruskal_then_permute(df$Q, df$cfg, B_perm = 2000)
out$letters
```

# From a result to the paper's numbers

`lcda_metrics()` turns a fitted result into the quantities the paper's tables
report. The default level is a tidy long table grouped by *scope*: the
partition (Q, community count, size distribution), the leaders (the NCE score
in both forms, leader degrees, coverage), the search (runtime, the best
iteration, trace dispersion, and how often the lexicographic tie-break on `H`
was actually decisive), and --- when a ground truth is supplied --- recovery.

```{r metrics}
m <- lcda_metrics(res_r)
subset(m, scope %in% c("partition", "search"))
```

Pass `truth =` for the recovery indices (NMI, ARI, Rand, VI, split-join). The
Zachary club's documented split is the natural reference here:

```{r metrics-truth}
truth <- c(rep(1, 17), rep(2, 17))
subset(lcda_metrics(res_r, truth = truth), scope == "recovery")
```

Baselines that carry no leaders are accepted and scored through exactly the
same code path, with their leaders derived as the top-eigenvector node of each
community and flagged as such, so a comparison table is one `rbind()` away:

```{r metrics-baseline}
subset(lcda_metrics(cluster_louvain(g), g, truth = truth), scope == "recovery")
```

The per-community and per-leader breakdowns answer the questions the summary
row cannot. Each community's `q_contribution` sums exactly to Q, and each
leader's `degree_pctile_in_community` is the statistic the paper's external
leader validation uses.

```{r metrics-levels}
lcda_metrics(res_r, level = "community")
lcda_metrics(res_r, level = "leader")
```

# The community-and-leader map

`plot()` on a fitted result draws the partition with the elected leaders
highlighted; the result carries the graph it was fitted on, so nothing else is
needed. `lcda_plot_communities()` does the whole pipeline (detect communities,
designate leaders, draw) from a bare graph in one call, and
`ggplot2::autoplot()` returns the same figure as a `ggplot` object.

```{r map, fig.width = 7, fig.height = 6}
set.seed(1)
plot(res_r)
```

# Community-conditioned NCE

The paper's NCE uses the *global* connection probability `k_v / n`. The
package also exposes a community-conditioned alternative
`nce_local_score()` measuring within-community connectivity entropy; the
two can differ materially (see the *Leader score* article).

```{r nce}
H_global <- nce_score(g, res_r$best$leaders)
H_local  <- nce_local_score(g, res_r$best$membership, res_r$best$leaders)
cat("H (global, paper):", round(H_global, 4),
    " | H (community-conditioned):", round(H_local, 4), "\n")
```

# Precomputed simulation datasets

The companion *Articles* render from precomputed results shipped under
`inst/extdata/`, so they never re-run a simulation. List them with:

```{r data}
lcda_data()
```

# References

- Ospina, R., Silva, G., Matos Junior, F. J., Leite, A., & Ochi, L. S.
  (2026). *A GRASP Framework for Community and Leader Detection in Complex
  Networks.* Preprint submitted to *Knowledge-Based Systems*.
- Feo, T. A., & Resende, M. G. C. (1995). Greedy Randomized Adaptive Search
  Procedures. *Journal of Global Optimization*, 6(2), 109--133.
  [doi:10.1007/BF01096763](https://doi.org/10.1007/BF01096763)
- Prais, M., & Ribeiro, C. C. (2000). Reactive GRASP. *INFORMS Journal on
  Computing*, 12(3), 164--176.
  [doi:10.1287/ijoc.12.3.164.12639](https://doi.org/10.1287/ijoc.12.3.164.12639)
