---
title: "Getting started with lcdaGRASP"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with lcdaGRASP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
```

# Overview

`lcdaGRASP` implements the four algorithms of Ospina, Silva, Matos Junior,
Leite and Ochi (2026) for **joint community-and-leader detection**:

- `lcda_grasp()` --- fixed-parameter GRASP, variants 1 (static centrality)
  and 2 (adaptive centrality recomputed each iteration).
- `lcda_gr()` --- Reactive GRASP, variants 1 and 2, with a bivariate
  self-tuning mechanism over a pool of `(alpha_c, alpha_s)` pairs.

The performance-critical kernels (incremental modularity, community
affinities, similarity vectors, the VNMI multi-improvement extension, and
power-iteration eigenvector centrality) are implemented in C++ via Rcpp.
The package follows a tidyverse style: snake_case throughout, `tibble`
outputs where tabular, `cli` for logging, and a `verbose` flag on every
substantive function.

# Quick start

```{r quick}
library(lcdaGRASP)
library(igraph)

g <- make_graph("Zachary")           # 34-node karate club
res <- lcda_grasp(g,
                  alpha_c = 0.1, alpha_s = 0.3, B = 30,
                  centrality = "eigen", similarity = "hpi",
                  seed = 1)
print(res)
```

Set `verbose = TRUE` for a `cli` progress bar and a one-line summary:

```{r verbose}
invisible(lcda_grasp(g, B = 10, seed = 1, verbose = TRUE))
```

# Inspecting the GRASP trajectory

```{r traj}
plot_grasp_trajectory(res)
```

# Switching to the Reactive variant

```{r reactive}
res_r <- lcda_gr(g, variant = 1, B = 100, seed = 1)
print(res_r)
plot_reactive_pk(res_r)
```

The print method reports the fraction of iterations in which the
lexicographic tie-break on `H` (Node-Connection Entropy) was decisive ---
iterations where `Q` was tied with the running best and `H` made the
difference. In typical networks this fraction is small; see the
*Reactive convergence* and *Pool sensitivity* articles.

# Statistical comparison

The paper uses a Kruskal--Wallis omnibus test followed by pairwise
permutation tests with Bonferroni correction. `kruskal_then_permute()`
returns a tibble of pairwise contrasts and a compact-letter display:

```{r kw}
set.seed(1)
df <- data.frame(
  Q = c(rnorm(20, 0.40, 0.01), rnorm(20, 0.42, 0.01), rnorm(20, 0.42, 0.01)),
  cfg = rep(c("a", "b", "c"), each = 20)
)
out <- kruskal_then_permute(df$Q, df$cfg, B_perm = 2000)
out$letters
```

# Community-conditioned NCE

The paper's NCE uses the *global* connection probability `k_v / n`. The
package also exposes a community-conditioned alternative
`nce_local_score()` measuring within-community connectivity entropy; the
two can differ materially (see the *Leader score* article).

```{r nce}
H_global <- nce_score(g, res_r$best$leaders)
H_local  <- nce_local_score(g, res_r$best$membership, res_r$best$leaders)
cat("H (global, paper):", round(H_global, 4),
    " | H (community-conditioned):", round(H_local, 4), "\n")
```

# Precomputed simulation datasets

The companion *Articles* render from precomputed results shipped under
`inst/extdata/`, so they never re-run a simulation. List them with:

```{r data}
lcda_data()
```

# References

- Ospina, R., Silva, G., Matos Junior, F. J., Leite, A., & Ochi, L. S.
  (2026). *A GRASP Framework for Community and Leader Detection in Complex
  Networks.* Preprint submitted to *Knowledge-Based Systems*.
- Feo, T. A., & Resende, M. G. C. (1995). Greedy Randomized Adaptive Search
  Procedures. *Journal of Global Optimization*, 6(2), 109--133.
  [doi:10.1007/BF01096763](https://doi.org/10.1007/BF01096763)
- Prais, M., & Ribeiro, C. C. (2000). Reactive GRASP. *INFORMS Journal on
  Computing*, 12(3), 164--176.
  [doi:10.1287/ijoc.12.3.164.12639](https://doi.org/10.1287/ijoc.12.3.164.12639)
