---
title: "Modularity degeneracy (Box)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modularity degeneracy (Box)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7.2, fig.height = 4.4)
library(lcdaGRASP)
d <- lcda_data("degeneracy"); have <- !is.null(d)
has_gg <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r nodata, echo = FALSE, results = "asis", eval = !have}
cat("> **Note.** Dataset not installed. Run `data-raw/run_all_data.R` and rebuild.\n")
```

# Every model is a simplification

Box's lens: modularity is an approximation, and one of its known pathologies
is *degeneracy* --- many structurally distinct partitions can have $Q$ within
$\varepsilon$ of the maximum [Good, de Montjoye & Clauset, 2010]. For a
multi-start metaheuristic this is double-edged: lots of optima to find, but the
single partition you report (and therefore the leaders you name) is somewhat
arbitrary. We audit how many **distinct near-best partitions** LCDA-GRASP
finds and their mean pairwise NMI (low NMI = high degeneracy).

```{r table, eval = have}
knitr::kable(
  dplyr::transmute(d$results, graph, epsilon,
                   Q_star = round(Q_star, 4),
                   runs_in_band = n_runs_in_band,
                   distinct_partitions = n_distinct_partitions,
                   mean_NMI = round(mean_pairwise_NMI, 3)),
  caption = "Distinct near-best partitions and their similarity, per network and tolerance.")
```

```{r plot, eval = have && has_gg}
library(ggplot2)
ggplot(d$results, aes(factor(epsilon), n_distinct_partitions, fill = graph)) +
  geom_col(position = "dodge") +
  labs(title = "Distinct near-best partitions within Q* - epsilon",
       x = expression(epsilon), y = "number of distinct partitions") +
  theme_minimal(base_size = 10)
```

**Reading.** Even at small $\varepsilon$, several distinct partitions coexist
in the top band on the larger/diffuse networks, and the mean pairwise NMI sits
below 1 --- so the reported leader-community configuration is *one of several*
near-equivalent ones. This does not invalidate the method, but it bounds the
interpretability claim about "named leaders": the names can shift across seeds.
The paper acknowledges the resolution limit but not this degeneracy; it is now
flagged in the manuscript's Limitations (see `artigo/CHANGES.md`).

# References

- Good, B. H., de Montjoye, Y.-A., & Clauset, A. (2010). Performance of
  modularity maximization in practical contexts. *Phys. Rev. E*, 81, 046106.
  [doi:10.1103/PhysRevE.81.046106](https://doi.org/10.1103/PhysRevE.81.046106)
