---
title: "The leader score: global vs community-conditioned NCE"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The leader score: global vs community-conditioned NCE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.2)
library(lcdaGRASP)
d <- lcda_data("nce_alternatives"); have <- !is.null(d)
has_gg <- requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("dplyr", quietly = TRUE)
```

```{r nodata, echo = FALSE, results = "asis", eval = !have}
cat("> **Note.** Dataset not installed. Run `data-raw/run_all_data.R` and rebuild.\n")
```

# Is the NCE measuring the right thing?

The paper's Node-Connection Entropy uses the **global** connection
probability $p_1(v) = k_v / n$. But a hub whose degree is spread across *all*
communities can maximise this score while being a poor *community* leader.
Since a score is a simplification, it is worth checking what it actually
rewards --- which motivates a **community-conditioned** alternative:

$$ I_{\text{local}}(v) = H_b\!\left(\frac{|\Gamma(v) \cap C(v)|}{|C(v)| - 1}\right), $$

exposed in the package as `nce_local_score()`. We measure how often the
designated leader **changes** when re-picked under the local score.

```{r table, eval = have}
d$results |>
  dplyr::group_by(graph) |>
  dplyr::summarise(
    communities       = round(mean(n_communities), 1),
    H_global          = round(mean(H_paper_global), 3),
    H_local           = round(mean(H_paper_local), 3),
    pct_leaders_changed = round(mean(pct_leaders_changed), 1),
    .groups = "drop") |>
  knitr::kable(caption = "Mean leader-score values and the fraction of leaders that change identity under the local NCE.")
```

```{r plot, eval = have && has_gg}
library(ggplot2)
ggplot(d$results, aes(graph, pct_leaders_changed, fill = graph)) +
  geom_boxplot(alpha = 0.6, show.legend = FALSE) +
  geom_hline(yintercept = 10, linetype = 2, colour = "grey40") +
  labs(title = "Leaders reassigned when switching global -> community-conditioned NCE",
       subtitle = "Dashed line = 10%: above it, the NCE choice materially changes who is a leader",
       x = NULL, y = "% of leaders changed") +
  theme_minimal(base_size = 10)
```

**Reading.** If the change fraction is non-trivial (above ~10%), the choice of
NCE definition is *not* an interpretive detail --- it materially decides who is
crowned the leader. Both scores are exported (`nce_score()` and
`nce_local_score()`) precisely so this comparison is part of the contribution
rather than buried.

# References

- Ospina, R., Silva, G., Matos Junior, F. J., Leite, A., & Ochi, L. S. (2026).
  *A GRASP Framework for Community and Leader Detection in Complex Networks.*
  Preprint.
- Akachar, E., Bougteb, Y., Ouhbi, B., & Frikh, B. (2025). LeaDCD:
  Leadership concept-based method for community detection. *Information
  Sciences*, 686, 121341.
  [doi:10.1016/j.ins.2024.121341](https://doi.org/10.1016/j.ins.2024.121341)
