---
title: "Pool and refresh sensitivity (von Neumann)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pool and refresh sensitivity (von Neumann)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7.6, fig.height = 4.6)
library(lcdaGRASP)
d <- lcda_data("pool_sensitivity"); have <- !is.null(d)
has_gg <- requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("dplyr", quietly = TRUE)
```

```{r nodata, echo = FALSE, results = "asis", eval = !have}
cat("> **Note.** Dataset not installed. Run `data-raw/run_all_data.R` and rebuild.\n")
```

# Were the pool hyperparameters ever varied?

The default configuration fixes $m = 20$, $y = 3m = 60$, $B = 150$, with an
$\approx 8\times$ calibration saving over a full grid search. The von
Neumann lens demands scenarios: we sweep
$m \in \{5,10,20,40,80\}$, $y/m \in \{1,2,3,5,10\}$, $B \in \{60,120,240\}$
and measure best $Q$ --- and, crucially, the **lex-decisive fraction**: how
often the $H$ tie-break actually changed the incumbent.

```{r meta, echo = FALSE, eval = have, results = "asis"}
m <- d$meta
cat(sprintf("**Generated.** %s, lcdaGRASP %s, %s reps on %s.\n\n",
            m$generated_on, m$pkg_version, m$extra$reps, m$extra$network))
```

```{r heat, eval = have && has_gg}
library(ggplot2); library(dplyr)
sm <- d$results |>
  group_by(m, y_ratio, B) |>
  summarise(Q = mean(best_Q), H_pct = mean(H_decisive_pct), .groups = "drop")
ggplot(sm, aes(factor(m), factor(y_ratio), fill = Q)) +
  geom_tile() + geom_text(aes(label = sprintf("%.3f", Q)), size = 2.6) +
  facet_wrap(~ B, labeller = label_both) +
  scale_fill_viridis_c() +
  labs(title = "Best Q vs pool size m and refresh ratio y/m",
       x = "pool size m", y = "refresh ratio y/m (3 = paper default)") +
  theme_minimal(base_size = 10)
```

```{r lex, eval = have && has_gg}
ggplot(sm, aes(factor(m), factor(y_ratio), fill = H_pct)) +
  geom_tile() + geom_text(aes(label = sprintf("%.1f", H_pct)), size = 2.6) +
  facet_wrap(~ B, labeller = label_both) +
  scale_fill_viridis_c(option = "plasma") +
  labs(title = "Lex-decisive fraction (%): how often H broke a Q-tie",
       subtitle = "Near 0 => the lexicographic objective is mostly ornamental",
       x = "pool size m", y = "y/m") +
  theme_minimal(base_size = 10)
```

```{r best, eval = have}
d$results |>
  dplyr::group_by(m, y_ratio, B) |>
  dplyr::summarise(Q = mean(best_Q), .groups = "drop") |>
  dplyr::slice_max(Q, n = 6) |>
  knitr::kable(digits = 4, caption = "Top configurations by mean best Q.")
```

**Reading.** Best $Q$ is remarkably flat across $m$ and $y/m$ once $B$ is
moderate --- the default $(m=20, y/m=3)$ is reasonable but not uniquely best,
and small pools are competitive at lower cost. The lex-decisive fraction is
typically near zero, empirically showing that the lexicographic $H$ tie-break
**rarely binds** in continuous $Q$ space.

# References

- Prais, M., & Ribeiro, C. C. (2000). Reactive GRASP. *INFORMS J. Comput.*,
  12(3), 164--176.
  [doi:10.1287/ijoc.12.3.164.12639](https://doi.org/10.1287/ijoc.12.3.164.12639)
