---
title: "Reactive update convergence"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reactive update convergence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7.2, fig.height = 4.4)
library(lcdaGRASP)
sm <- lcda_data("prop6_summary"); tr <- lcda_data("prop6_trajectories")
have <- !is.null(sm) && !is.null(tr)
has_gg <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r nodata, echo = FALSE, results = "asis", eval = !have}
cat("> **Note.** Dataset not installed. Run `data-raw/run_all_data.R` and rebuild.\n")
```

# From an asymptotic theorem to a finite-budget question

Proposition 6 proves that the Reactive update concentrates the selection
probabilities $p_k$ on the best parameter pair as $B \to \infty$. The von
Neumann lens asks the *finite-budget* questions the proof does not: **how
fast** does concentration happen, and **is the empirical winner the true
winner** at the $B = 150$ the paper actually runs? We answer both by tracking
the Shannon entropy $H(p_k)$ for $B \in \{150, 300, 600, 1200\}$.

```{r entropy, eval = have && has_gg}
library(ggplot2)
ggplot(tr$results, aes(iter, H_pk, colour = factor(B_max))) +
  geom_line(alpha = 0.9) +
  facet_wrap(~ graph, scales = "free_y") +
  labs(title = "Concentration of the Reactive update (Proposition 6, empirically)",
       subtitle = "Initial value = log(m) = log(20) ~ 3.0; lower = more concentrated",
       x = "GRASP iteration", y = "Shannon entropy of p_k", colour = "B") +
  theme_minimal(base_size = 10)
```

```{r table, eval = have}
sm$results |>
  dplyr::transmute(graph, B,
                   H_initial = round(H_p_initial, 3),
                   H_final   = round(H_p_final, 3),
                   H_max     = round(H_max, 3),
                   rank_corr = round(rank_corr_p_mu, 3),
                   best_Q    = round(best_Q, 4)) |>
  knitr::kable(caption = "Entropy drop and rank correlation between final p_k and mean performance mu_k.")
```

**Reading.** Entropy falls from $\log 20 \approx 3.0$ toward a low plateau
within the first few refresh blocks, and the rank correlation between $p_k$
and $\mu_k$ is high --- the empirical winner matches the best-mean pair. The
practical implication: the paper's $B = 150$ is enough for the *concentration*
to occur, but (see the *Pool sensitivity* article) concentration is not the
same as solution quality.

# References

- Prais, M., & Ribeiro, C. C. (2000). Reactive GRASP. *INFORMS Journal on
  Computing*, 12(3), 164--176.
  [doi:10.1287/ijoc.12.3.164.12639](https://doi.org/10.1287/ijoc.12.3.164.12639)
