---
title: "Robustness across community-strength regimes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Robustness across community-strength regimes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7.2, fig.height = 4.4)
library(lcdaGRASP)
d <- lcda_data("lfr_robustness"); have <- !is.null(d)
has_gg <- requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("dplyr", quietly = TRUE)
```

```{r nodata, echo = FALSE, results = "asis", eval = !have}
cat("> **Note.** Dataset not installed. Run `data-raw/run_all_data.R` and rebuild.\n")
```

# The coverage gap

The paper evaluates on five well-behaved networks; one should not conclude
only from where there is data. We sweep the **mixing
parameter** $\mu \in [0.1, 0.8]$ on **canonical LFR** networks
(Lancichinetti--Fortunato--Radicchi 2008) --- from strong community structure
($\mu \to 0$) into the regime where the planted communities dissolve --- and
measure $Q$, the adjusted Rand index (ARI) **and** normalized mutual
information (NMI) against the planted partition. $Q$ alone cannot tell you
whether the *recovered* partition is correct; ARI/NMI can.

```{r meta, echo = FALSE, eval = have, results = "asis"}
m <- d$meta
cat(sprintf("**Generated.** %s, lcdaGRASP %s, %s reps, n=%s.\n\n",
            m$generated_on, m$pkg_version, m$extra$reps, m$extra$n))
if (!is.null(m$extra$generator)) cat("- Generator: ", m$extra$generator, "\n")
cat(paste0("- ", m$limitations, collapse = "\n"), "\n")
```

```{r summ, eval = have}
quality <- d$results$quality
sm <- quality |>
  dplyr::group_by(mu, method) |>
  dplyr::summarise(NMI = mean(NMI), NMI_sd = sd(NMI), ARI = mean(ARI),
                   Q = mean(Q), .groups = "drop")
```

```{r plot-nmi, eval = have && has_gg}
library(ggplot2)
ggplot(sm, aes(mu, NMI, colour = method, fill = method)) +
  geom_ribbon(aes(ymin = NMI - NMI_sd, ymax = NMI + NMI_sd), alpha = 0.12, colour = NA) +
  geom_line(linewidth = 0.9) + geom_point(size = 1.3) +
  labs(title = "Recovery of the planted partition (NMI vs ground truth)",
       subtitle = "Canonical LFR; higher is better",
       x = expression(mu~"(mixing parameter)"), y = "NMI") +
  theme_minimal(base_size = 10)
```

```{r table, eval = have && requireNamespace("tidyr", quietly = TRUE)}
sm |>
  dplyr::filter(mu %in% c(0.2, 0.4, 0.6)) |>
  dplyr::transmute(mu, method, NMI = round(NMI, 3)) |>
  tidyr::pivot_wider(names_from = mu, values_from = NMI) |>
  knitr::kable(caption = "NMI vs the planted partition at low/medium/high mixing.")
```

**Reading (honest).** On canonical LFR the proposed methods recover well at low
mixing but **trail** Leiden/Louvain --- and especially **ECG** (ensemble
clustering), the strongest recovery baseline --- across the sweep. This is the
honest counterpart to the five hand-picked benchmarks: LCDA-GRASP/GR is a
*competitive, not leading* community detector. The ensemble-consensus variant
`lcda_ecg()` closes this gap (see *Reproducing the paper's benchmark numbers*
and the package's `lcda_ecg` documentation).

## Leader utility across community strength

The pool also lets us ask whether the LCDA leaders are useful *seeds*: we
compare the expected Independent-Cascade spread from the detected leaders
against an equal-size top-degree seed set, as a function of $\mu$.

```{r leader, eval = have && has_gg && !is.null(d$results$leader_utility) && nrow(d$results$leader_utility) > 0}
u <- d$results$leader_utility |>
  dplyr::group_by(mu) |>
  dplyr::summarise(adv_vs_degree = mean(adv_vs_degree),
                   se = sd(adv_vs_degree) / sqrt(dplyr::n()),
                   adv_vs_random = mean(adv_vs_random), .groups = "drop")
ggplot(u, aes(mu, adv_vs_degree)) +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey50") +
  geom_ribbon(aes(ymin = adv_vs_degree - se, ymax = adv_vs_degree + se), alpha = 0.15) +
  geom_line(linewidth = 0.9) + geom_point(size = 1.3) +
  labs(title = "Leader spreading advantage vs a top-degree seed set",
       subtitle = "Above zero: LCDA leaders out-spread top-degree of the same size",
       x = expression(mu), y = "IC-spread advantage (LCDA - top-degree)") +
  theme_minimal(base_size = 10)
```

**Reading.** The leaders always beat a random seed set, but they do **not**
out-spread a top-degree set on LFR; any edge is confined to the strongest-
community regime and inverts as $\mu$ grows. The leader contribution is the
*joint, community-grounded* designation --- not superior raw spreading.

# References

- Lancichinetti, A., Fortunato, S., & Radicchi, F. (2008). Benchmark graphs
  for testing community detection algorithms. *Phys. Rev. E*, 78, 046110.
  [doi:10.1103/PhysRevE.78.046110](https://doi.org/10.1103/PhysRevE.78.046110)
- Poulin, V., & Théberge, F. (2019). Ensemble clustering for graphs.
  *Applied Network Science*, 4, 51.
  [doi:10.1007/s41109-019-0162-z](https://doi.org/10.1007/s41109-019-0162-z)
