---
title: "Robustness and honest limits: recovery, distributions, and degeneracy"
output: rmarkdown::html_vignette
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7.4, fig.height = 4.6)
library(lcdaGRASP)

d_lfr <- lcda_data("lfr_robustness")
d_eda <- lcda_data("eda_replicates")
d_deg <- lcda_data("degeneracy")

have_lfr <- !is.null(d_lfr)
have_eda <- !is.null(d_eda)
have_deg <- !is.null(d_deg)

has_gg    <- requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("dplyr", quietly = TRUE)
has_tidyr <- requireNamespace("tidyr", quietly = TRUE)
```

The paper evaluates LCDA-GRASP/GR on five well-behaved benchmark networks. One
should not conclude only from where there is data: those five networks have
strong, clean community structure, and modularity is only an approximation of
"good partition". This article stresses the method along three axes that the
headline tables do not show, and reports what it finds without embellishment.

The thread running through all three is a single discipline: **do not trust the
mean alone.** Recovery degrades as communities dissolve; the per-replicate
spread is wide where tables print only a point estimate; and several distinct
"near-best" partitions coexist, so the one you report is somewhat arbitrary. The
honest reading of the package is that LCDA-GRASP/GR is a *competitive, not
leading* community detector, and its leader designation is valuable as a
*joint, community-grounded* output rather than as a superior raw spreader.

# 1. Recovery across community-strength regimes

We sweep the **mixing parameter** $\mu \in [0.1, 0.8]$ on **canonical LFR**
networks (Lancichinetti--Fortunato--Radicchi 2008) --- from strong community
structure ($\mu \to 0$) into the regime where the planted communities dissolve
--- and measure $Q$, the adjusted Rand index (ARI) **and** normalized mutual
information (NMI) against the planted partition. $Q$ alone cannot tell you
whether the *recovered* partition is correct; ARI/NMI can.

```{r lfr-nodata, echo = FALSE, results = "asis", eval = !have_lfr}
cat("> **Note.** The `lfr_robustness` dataset is not installed. Run `data-raw/run_all_data.R` and rebuild.\n")
```

```{r lfr-meta, echo = FALSE, eval = have_lfr, results = "asis"}
m <- d_lfr$meta
cat(sprintf("**Generated.** %s, lcdaGRASP %s, %s reps, n=%s.\n\n",
            m$generated_on, m$pkg_version, m$extra$reps, m$extra$n))
if (!is.null(m$extra$generator)) cat("- Generator: ", m$extra$generator, "\n")
cat(paste0("- ", m$limitations, collapse = "\n"), "\n")
```

```{r lfr-summ, eval = have_lfr && has_gg}
library(ggplot2); library(dplyr)
quality <- d_lfr$results$quality
sm <- quality |>
  dplyr::group_by(mu, method) |>
  # Dispersion first: inside summarise() a later expression sees the column it
  # has just created, so sd(NMI) after NMI = mean(NMI) would receive a scalar
  # and return NA (silently dropping the ribbon below).
  dplyr::summarise(NMI_sd = sd(NMI), NMI = mean(NMI), ARI = mean(ARI),
                   Q = mean(Q), .groups = "drop")
```

```{r lfr-plot-nmi, eval = have_lfr && has_gg}
ggplot(sm, aes(mu, NMI, colour = method, fill = method)) +
  geom_ribbon(aes(ymin = NMI - NMI_sd, ymax = NMI + NMI_sd), alpha = 0.12, colour = NA) +
  geom_line(linewidth = 0.9) + geom_point(size = 1.3) +
  labs(title = "Recovery of the planted partition (NMI vs ground truth)",
       subtitle = "Canonical LFR; shaded band = ±1 sd over replicates; higher is better",
       x = expression(mu~"(mixing parameter)"), y = "NMI") +
  theme_minimal(base_size = 10)
```

```{r lfr-table, eval = have_lfr && has_gg && has_tidyr}
sm |>
  dplyr::filter(mu %in% c(0.2, 0.4, 0.6)) |>
  dplyr::transmute(mu, method, NMI = round(NMI, 3)) |>
  tidyr::pivot_wider(names_from = mu, values_from = NMI) |>
  knitr::kable(caption = "NMI vs the planted partition at low/medium/high mixing.")
```

On canonical LFR the proposed methods recover well at low mixing but **trail**
Leiden/Louvain --- and especially **ECG** (ensemble clustering, Poulin &
Théberge 2019), the strongest recovery baseline --- across the sweep. The
shaded band already makes the discipline concrete: the per-replicate sd is
non-trivial, so two methods whose means cross may be indistinguishable at a
given $\mu$. The ensemble-consensus variant `lcda_ecg()` closes the recovery
gap (see *Reproducing the paper's benchmark numbers* and the `lcda_ecg`
documentation).

## Leader utility across community strength

The same pool lets us ask whether the LCDA leaders are useful *seeds*: we
compare the expected Independent-Cascade spread from the detected leaders
against an equal-size top-degree seed set, as a function of $\mu$.

```{r lfr-leader, eval = have_lfr && has_gg && !is.null(d_lfr$results$leader_utility) && nrow(d_lfr$results$leader_utility) > 0}
u <- d_lfr$results$leader_utility |>
  dplyr::group_by(mu) |>
  # Standard error first (see the note above: a self-referencing summarise()
  # would hand sd() the mean it has just computed and yield NA).
  dplyr::summarise(se = sd(adv_vs_degree) / sqrt(dplyr::n()),
                   adv_vs_degree = mean(adv_vs_degree),
                   adv_vs_random = mean(adv_vs_random), .groups = "drop")
ggplot(u, aes(mu, adv_vs_degree)) +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey50") +
  geom_ribbon(aes(ymin = adv_vs_degree - se, ymax = adv_vs_degree + se), alpha = 0.15) +
  geom_line(linewidth = 0.9) + geom_point(size = 1.3) +
  labs(title = "Leader spreading advantage vs a top-degree seed set",
       subtitle = "Above zero: LCDA leaders out-spread top-degree of the same size",
       x = expression(mu), y = "IC-spread advantage (LCDA − top-degree)") +
  theme_minimal(base_size = 10)
```

The leaders always beat a random seed set, but they do **not** out-spread a
top-degree set on LFR. The $\pm 1$ standard-error band makes the reading
precise: at the strongest-community regime the difference is indistinguishable
from zero (the band still covers the dashed line), and it turns clearly
negative as $\mu$ grows. The leader contribution is the *joint,
community-grounded* designation --- not superior raw spreading.

# 2. Distributions of Q and H, not just means

Where the paper reports means and significance letters, here we draw the
**boxplots, ECDFs and (Q, H) scatter** that those tables hide, over
replications per (graph $\times$ algorithm $\times$ centrality $\times$
similarity).

```{r eda-nodata, echo = FALSE, results = "asis", eval = !have_eda}
cat("> **Note.** The `eda_replicates` dataset is not installed. Run `data-raw/run_all_data.R` and rebuild.\n")
```

```{r eda-prep, eval = have_eda && has_gg}
res <- d_eda$results |> dplyr::mutate(config = paste(centrality, similarity, sep = "/"))
```

```{r eda-box, eval = have_eda && has_gg}
# Q on x directly (coord_flip would silently disable the facet free scale).
ggplot(res, aes(Q, config, fill = algorithm)) +
  geom_boxplot(outlier.size = 0.5, alpha = 0.65, orientation = "y") +
  facet_wrap(~ graph, scales = "free_x") +
  labs(title = "Best-Q by configuration (each box = replications)",
       subtitle = "Means alone hide this picture", y = NULL, x = "best Q") +
  theme_minimal(base_size = 9)
```

```{r eda-ecdf, eval = have_eda && has_gg}
ggplot(res, aes(Q, colour = config)) +
  stat_ecdf(geom = "step", linewidth = 0.7) +
  facet_wrap(~ graph, scales = "free_x") +
  labs(title = "ECDF of best-Q by configuration", x = "Q", y = "F(Q)") +
  theme_minimal(base_size = 9) + theme(legend.position = "bottom")
```

```{r eda-qh, eval = have_eda && has_gg}
# Common Q scale (free_y), so the near-fixed Q reads as narrow vertical clouds
# instead of being auto-zoomed apart per panel.
ggplot(res, aes(Q, H, colour = algorithm, shape = similarity)) +
  geom_point(alpha = 0.6, size = 1.4) +
  facet_wrap(~ graph, scales = "free_y") +
  labs(title = "Joint (Q, H) across replicates",
       subtitle = "H varies within a narrow Q band: the near-tie regime where a lexicographic Q-then-H order could act") +
  theme_minimal(base_size = 9)
```

```{r eda-winners, eval = have_eda && has_gg}
d_eda$results |>
  dplyr::group_by(graph, centrality, similarity) |>
  dplyr::summarise(Q = mean(Q), .groups = "drop_last") |>
  dplyr::slice_max(Q, n = 1) |>
  dplyr::ungroup() |>
  knitr::kable(digits = 4, caption = "Best centrality/similarity per network (mean Q).")
```

HPI + eigenvector is the modal winner, matching the paper's recommendation ---
but the boxplots show the margins are often within noise, and the (Q, H) clouds
are nearly vertical, i.e. $H$ varies at essentially fixed $Q$. That is exactly
the near-tie situation in which a lexicographic (Q, then H) objective could in
principle bite. In practice it rarely does: $H$ seldom breaks a $Q$-tie, as
quantified in the *Pool sensitivity* article. The picture reinforces the same
discipline as the LFR sweep --- report the distribution, because the gap between
configurations is frequently smaller than the within-configuration spread.

# 3. Modularity degeneracy and partition multiplicity

Modularity is an approximation, and one of its known pathologies is
*degeneracy*: many structurally distinct partitions can have $Q$ within
$\varepsilon$ of the maximum (Good, de Montjoye & Clauset 2010). For a
multi-start metaheuristic this is double-edged --- lots of optima to find, but
the single partition you report (and therefore the leaders you name) is somewhat
arbitrary. We audit how many **distinct near-best partitions** LCDA-GRASP finds
and their mean pairwise NMI (low NMI = high degeneracy).

```{r deg-nodata, echo = FALSE, results = "asis", eval = !have_deg}
cat("> **Note.** The `degeneracy` dataset is not installed. Run `data-raw/run_all_data.R` and rebuild.\n")
```

```{r deg-table, eval = have_deg && has_gg}
knitr::kable(
  dplyr::transmute(d_deg$results, graph, epsilon,
                   Q_star = round(Q_star, 4),
                   runs_in_band = n_runs_in_band,
                   distinct_partitions = n_distinct_partitions,
                   mean_NMI = round(mean_pairwise_NMI, 3)),
  caption = "Distinct near-best partitions and their similarity, per network and tolerance.")
```

```{r deg-plot, eval = have_deg && has_gg}
ggplot(d_deg$results, aes(factor(epsilon), n_distinct_partitions, fill = graph)) +
  geom_col(position = "dodge") +
  labs(title = "Distinct near-best partitions within Q* − epsilon",
       x = expression(epsilon), y = "number of distinct partitions") +
  theme_minimal(base_size = 10)
```

Even at small $\varepsilon$, several distinct partitions coexist in the top band
on the larger/diffuse networks, and the mean pairwise NMI sits below 1 --- so
the reported leader-community configuration is *one of several* near-equivalent
ones. This does not invalidate the method, but it bounds the interpretability
claim about "named leaders": the names can shift across seeds. Beyond the
well-known resolution limit, this near-optimum degeneracy is a genuine caveat
for any modularity-maximising leader assignment and is worth reporting alongside
it.

# Taking the three together

The same lesson recurs at every scale of the analysis. As communities weaken,
recovery falls and LCDA trails the strongest baselines while remaining
competitive and matching ECG-style ensembles; across replicates the
configuration margins are often inside the noise; and at the top of the
$Q$-landscape several partitions are statistically interchangeable, so the
"named leaders" are not unique. None of this sinks the method --- but it is the
honest envelope around the headline numbers, and it is the reason the package
reports distributions, leader-utility comparisons, and degeneracy counts rather
than point estimates alone.

# Reproducibility and data provenance

Every figure here is read from datasets shipped with the package; nothing is
simulated at build time. Each records the package version and generation date
and ships with a SHA-256 checksum (matching `inst/extdata/SHA256SUMS`):

```{r provenance}
do.call(rbind, lapply(c("lfr_robustness", "eda_replicates", "degeneracy"),
                      lcda_provenance))
```

Regenerate with `data-raw/30_lfr_robustness.R`, `data-raw/70_eda.R` and
`data-raw/60_degeneracy.R` (each fixes a single documented seed; the LFR sweep
uses $n = 500$ with 10 graphs per mixing level).

**Scope.** All synthetic results are at $n = 500$ (size effects are studied
separately in the large-scale evaluation); recovery is reported as a mean with
dispersion over the replicate graphs, never as a single number.

# References

- Good, B. H., de Montjoye, Y.-A., & Clauset, A. (2010). Performance of
  modularity maximization in practical contexts. *Phys. Rev. E*, 81, 046106.
  [doi:10.1103/PhysRevE.81.046106](https://doi.org/10.1103/PhysRevE.81.046106)
- Lancichinetti, A., Fortunato, S., & Radicchi, F. (2008). Benchmark graphs
  for testing community detection algorithms. *Phys. Rev. E*, 78, 046110.
  [doi:10.1103/PhysRevE.78.046110](https://doi.org/10.1103/PhysRevE.78.046110)
- Newman, M. E. J., & Girvan, M. (2004). Finding and evaluating community
  structure in networks. *Phys. Rev. E*, 69, 026113.
  [doi:10.1103/PhysRevE.69.026113](https://doi.org/10.1103/PhysRevE.69.026113)
- Poulin, V., & Théberge, F. (2019). Ensemble clustering for graphs.
  *Applied Network Science*, 4, 51.
  [doi:10.1007/s41109-019-0162-z](https://doi.org/10.1007/s41109-019-0162-z)
```
